Measuring Perceptions and Preferences for Meritocracy

### 2021-09-22 paper.Rmd
ajustes grammar

---
title: "Measuring Perceptions and Preferences for Meritocracy"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: "Economic and social inequalities have generated growing concerns and crises across contemporary societies. One of the mechanisms proposed by social sciences to explain the persistence of inequality is the belief in meritocracy, which would legitimize economic disparities based on differences in effort and talent. Despite its wide use as a concept, empirical research on meritocracy is relatively novel and characterized by diverse conceptualization and measures that make the findings and their interpretation rather inconsistent. Most of the studies in the area have relied upon secondary data to operationalize meritocracy, with a wide variation in the use and interpretation of the same survey items. Taking into account the extant literature that uses measures of meritocracy, this article identifies a series of drawbacks and inconsistencies within and between studies regarding conceptualization and operationalization. Based on this critical analysis, we propose a conceptual framework for studying perceptions and preferences for meritocracy and non-meritocracy, which is then tested through confirmatory analysis using ISSP (International Social Survey Programme) data as well as a novel survey designed specifically with this purpose (N=2,141). Our results support the conceptual framework and its operationalizarion, although with a better fit for the proposed scale than for the ISSP survey. Our discussion highlights the importance of considering different dimensions in order to advance in the study of meritocracy."
output:
  bookdown::pdf_document2:
     template: null
     toc: false
     number_sections: false
  bookdown::html_document2:
    number_sections: false
  bookdown::word_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/meritocracy.bib
csl: input/bib/apa7-singlespace.csl  
editor_options:
  chunk_output_type: console
geometry: margin=0.78in
header-includes:
  # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel}
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage[figurename=Fig.]{caption}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{titlesec}
  - \titleformat{\paragraph}
    {\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
    \titlespacing*{\paragraph}
    {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
always_allow_html: true # for word output
fontsize: 10pt
---

@@

\newpage

@@

Economic inequality and income concentration have become topics of growing concern over the last years, acquiring even more salience in times of pandemic and economic crises. This has led to a series of social movements in different societies as well as diverse critical analyses regarding the development of capitalism and its consequences [@streeck_politics_2014; @piketty_capital_2014]. In this context, the study of views, preferences and perceptions of inequality has acquired relevance in the social sciences, in topics such as redistributive preferences [@alesina_fairness_2005; @dimick_models_2018], the legitimization of economic inequality [@schroder_income_2017] and the functioning of meritocracy [@duru-bellat_who_2012; @mijs_paradox_2019; @reynolds_perceptions_2014; @atria_economic_2020]. Within this context, the ideal of meritocracy has been strongly challenged as an unfulfilled promise of modern societies that allows the perpetuation of social inequalities [@sandel_tyranny_2020; @goldthorpe_myth_2003].

In general, meritocracy is defined as a system of distribution of resources and rewards based on individual merit, which in its original conception is a combination of talent and effort [@young_rise_1962]. This traditional conception of merit places in a secondary position the possible interference of structural or non-meritocratic factors, such as inheritance, personal contacts, and luck [@breen_class_1999; @saunders_might_1995; @yair_meritocracy_2007; @land_we_2006; @young_rise_1962]. Social psychology and sociology have studied the characteristics and consequences of beliefs in meritocracy, under the general hypothesis that a greater belief in meritocracy emphasizes the role of individual over structural factors in personal achievements, leading to greater legitimization of inequalities [@preminger_meritocracy_2020; @trump_when_2020; @hadjar_meritokratie_2008; @madeira_primes_2019]. Such research has raised criticism of meritocracy as a moral standard of distribution given the preponderant weight of non-meritocratic elements upon individual status and social mobility[@sandel_tyranny_2020; @witteveen_reconsidering_2020; @arrow_meritocracy_2000; @goldthorpe_myth_2003; @markovits_meritocracy_2019; @khan_privilege_2013].

Due to the role that meritocratic beliefs play in the justification of individual achievement (or failure) in contemporary societies [@hadjar_meritokratie_2008, @markovits_meritocracy_2019; @sandel_tyranny_2020], multiple studies have evaluated the relationship between meritocratic beliefs and personal and/or contextual characteristics. For example, some studies have linked meritocracy to the reinforcement of socio-economic, gender, and ethnic stereotypes [@madeira_primes_2019; @girerd_neoliberalism_2020; @girerd_neoliberalism_2020], as well as the effects of meritocratic beliefs in educational [@generett_stories_2020; @owens_engines_2020] and organizational contexts [@perez_advancing_2020; @aiello_new_2019]. Most of such studies so far have used indicators from existing standard social surveys, such as the International Social Survey Programme (ISSP), to measure meritocracy. However, as we will show later, the concepts used as well as the instruments for  measuring meritocracy vary extensively among studies. In many cases, similar phenomena are associated with different indicators, and conversely, often different phenomena are measured with similar indicators. Such trends not only limits the comparability of studies, but also the ability to understanding the effects of meritocratic beliefs across different disciplines and lines of research.

@@

In the following we discuss four critical aspects in studies conceptualizing and measuring meritocracy and that are the basis of the conceptual and measurement proposal.

@@

One recent definition of meritocracy by @mijs_paradox_2019 is the following: "when I discuss meritocracy beliefs, I am referring to citizens’ belief in the importance of hard work relative to structural factors." [@mijs_paradox_2019, pg.9]. In the subsequent operationalization, this is associated with the following question and indicator: "how important you think it is for getting ahead in life: (a) hard work", scored from 1 to 5 on a likert scale. The assumptions behind such definition are worth discussing in light of the conceptual meaning of meritocracy and its possibilities of operationalization.

The item used by @mijs_paradox_2019 is part of an items' battery present in several international surveys, usually called "reasons to get ahead". This battery displays a series of indicators related to what people consider important to get ahead in life: hard work, education, ambition, a wealthy family, the right connections, religion, race, and gender. Therefore, when considering only one of the items, it means that other aspects that could be associated with talent, such as education, would not be deemed as meritocratic. As he points out: "Hard work is arguably the most meritocratic part of Michael Young’s equation: ‘Merit = Intelligence + Effort’, for the simple fact that intelligence itself is influenced by a non-meritocratic factor: who your parents happen to be" (p.5).

In this conceptual and measurement approach of meritocratic beliefs, we can observe a couple of strong conceptual assumptions: a) effort would not depend on parental influence, and b) talent (as innate ability) is not meritocratic (contrary to Michael Youngs' original conceptualization). This conceptual and measurement-based assumption is found in other studies, that also assume effort to be the main and only aspect of meritocracy [@girerd_neoliberalism_2020; @bubak_perceptions_2019], raising the question: Is effort the only dimension behind the concept of meritocracy? The question of whether talent (as inteligence and ability) is or is not considered meritocratic is certainly an interesting topic to discuss from a philosophical point of view, but for those working on empirical studies it should face empirical scrutiny. Besides, the consideration of talent as part of meritocracy certainly opens some interesting avenues of research, for instance in studies showing that for the elites meritocracy is more related to talent, whereas effort is something more characteristic of meritocracy of the middle and low classes [@atria_economic_2020].

@@

Several approaches to the empirical study of meritocracy based on public opinion surveys refer to the concept of _beliefs_, but with wide differences in meanings and operationalizations. To illustrate this point, lets consider the most commonly used items for operationalizing meritocratic beliefs in survey research (as ISSP), which is the "reasons to get ahead" battery. It consists in a series of items asking "how important you think it is for getting ahead in life" and then lists several factors, such as effort, education, parental wealth, and contacts. Another version of this same battery used in other surveys - sometimes together with the previous one - asking about "how important you think it _should_ be ...", and then listing the same concepts.  Therefore, the question raised here is: Which one of both is a "belief": what _is_ or what it _should_ be?

The term belief has an ambiguous character in the literature, conceived as "idea-elements" by @converse_nature_1964 or "considerations" by @zaller_nature_1992. As @kluegel_beliefs_1986 pointed out about the scope of beliefs: "This usage encompasses such more specific social-psychological concepts as values, perceptions, and attitudes" (p.30). Therefore, beliefs cover almost anything related to subjective factors. To this regard, a relevant distinction in the field of inequality beliefs was made by @janmaat_subjective_2013: "Perceptions refer to subjective estimates of existing inequality (i.e. thoughts about what is). Beliefs are here defined as normative ideas about just inequality (i.e. thoughts about what should be)"(p.359). Several papers dealing with meritocracy use the term beliefs (i.e. what should be), while actually referring to perceptions (i.e. what is). This occurs for instance in @reynolds_perceptions_2014, in which the term beliefs is used to talk about what @janmaat_subjective_2013 refer as perceptions, whereas other authors use general terms such as attitudes [@kunovich_systems_2007]. The first attempt to shed light on this issue in meritocracy research was made by @duru-bellat_who_2012, who used the question "how important should the number of years spent in education and training be in deciding how much money people ought to earn?" as a proxy for "desired" meritocracy (normative beliefs). They then determined "perceived" meritocracy, using the questions: “Would you say that in your country, people are rewarded for their efforts?” and “... people are rewarded for their skills?”.

Is the belief in meritocracy a perception or a preference with normative meaning? In order to expand the analytical conceptual framework, we believe that both dimensions should be included as proposed by @duru-bellat_who_2012. This opens up the possibility of analyzing whether perceptions and preferences are actually related (i.e. have a high correlation), or whether they are independent aspects of the same phenomenon. As @sonhing_merit_2011 have pointed out, "People can believe that outcomes ought to be distributed on the basis of merit and yet vary in their perceptions of whether this is how society currently operates" (p. 435). In other words, normative beliefs should be considered while taking perceptions into account: a strong normative belief in meritocracy may mean something totally different to someone perceiving high meritocracy, when compared to someone perceiving low meritocracy. To avoid the confusion generated by the term "belief", we propose the terms meritocratic preferences ("what should be"), and meritocratic perceptions ("what is"), as they better reflect the two facets of meritocracy under scrutiny [@castillo_meritocracia_2019].

@@

Some research in meritocracy considers aspects usually opposed to effort and talent for personal achievement, as for instance the use of personal advantages (as contacts or having a wealthy family) to get ahead in life. For instance, @kunovich_systems_2007 used an items’ battery listing a number of factors in relation to “How important each should be in deciding pay...” (as @duru-bellat_who_2012 for desired meritocracy). They consider factors such as education and responsibility as meritocratic, giving them a value of 1 if considered “essential” in the scale response, whereas factors such as having a family or children are valued 1 when rated as "not important at all" (i.e. reverse coded). The assumption behind this approach is that rejecting a supposed non-meritocratic aspect (as having family and/or children) implies a stronger belief in meritocracy. A similar approach of reverse-coding non-meritocratic items was taken by @newman_false_2015, using the same principles applied in the "Preference for the Merit Principle Scale" [@davey_preference_1999].

@@

The columns "Perceptions" and "Preferences" represent the distinction between these two concepts, usually confused under the label "beliefs" [@castillo_meritocracia_2019]. Perceptions refer to the extent to which people observe that meritocracy functions or applies in their society, which relates to items such as "Hard work is important to get ahead in society". Preferences refer to normative expectations that are usually linked to a "should" expression (e.g. whether hard work should be related to payment). The rows in Figure \@ref(fig:merit-model) consider the distinction between meritocratic and non-meritocratic dimensions [@reynolds_perceptions_2014], usually treated as different ends of the same continuum in previous research. Non-meritocratic elements refer to the use of resources such as personal contacts or family advantages to get ahead in life.

@@

The analysis is organized in three studies. Study 1 serves as a background to our instrument proposal by attempting to operationalize the meritocracy dimensions with available secondary data from the inequality module of the ISSP. Although we are aware that the available indicators in this dataset do not allow a precise operationalization of the conceptual model, we know that it is the most widely used data in comparative meritocracy studies (as well as overall subjective inequality) and therefore it is certainly useful to look at its potentials and limitations to this regard. Study 2 is the core of this paper and consists of the proposal of a new scale for measuring meritocratic perceptions and preferences. Finally, in Study 3 we perform additional validity analysis of the meritocracy scale.

@@

The data cooresponds to the last available wave of the social inequality module from the ISSP, which is the most specialized international comparative survey in perceptions, attitudes and beliefs about inequality related issues. This wave corresponds to the year 2009 and covers attitudes towards a series of topics dealing with social inequality across 41 countries and 56021 individuals. Although there is data available from this module for previous waves (1987, 1992 and 1999), unfortunately there are several variables that are important for this research that were not included, particularly in 1987 and 1999, reason why we only use the 2009 wave (during the development of this research the 2019 wave became available, but so far only with a restricted set of countries).

@@

- Preferences for meritocracy-non meritocracy: for the operationalization of normative preferences we used a list of items related to reasons for pay battery. The ISSP question was: _In deciding how much people ought to earn, how important should each of these things be, in your opinion_, rated in the same _essential-non important at all_ scale (1 to 5) as the questions for meritocratic perceptions.

@@

The estimation was performed using Confirmatory Factor Analysis (CFA). CFA was conducted using the `lavaan` R package (version 0.6-3; @rosseel_lavaan_2012), with diagonally weighted least squares (DWLS) estimation due to the items’ ordinal level of measurement [@kline_principles_2016; @rosseel_lavaan_2012]. As recommended by @brown_confirmatory_2015, we assessed model fit by jointly considering the comparative fit index and Tucker-Lewis Index (CFI and TLI; acceptable fit > 0.95), Root of the average squared residual approximation (RMSEA; acceptable fit < 0.08),  Chi-square: (p-value; acceptable fit > 0.05, and Chi-square ratio > 3).

@@

```{r desc-issp, eval=FALSE, include=FALSE, results='asis'}

  load(file = "input/data/proc/data_s1.RData")
  if (!require("pacman")) install.packages("pacman")
  pacman::p_load(dplyr, psych, knitr, kableExtra)
  table_format = if(knitr::is_html_output()) { #conditional instructions for kable
    "html" #if html set "html" in format
  } else if(knitr::is_latex_output()) {
    "latex"#if latex set "latex" in format
  }

  data_s1 %>%
    dplyr::select(hwork,ambition,wealthy,pareduc,
                  race,gender,people,polcone,welljob,hardjob,family,child) %>%
    psych::describe() %>%
    as.data.frame() %>%
    dplyr::select("Mean"=mean,"SD"=sd,"Min"=min,"Max"=max) %>%
    round(2) ->desc.issp

  row.names(desc.issp) <- c(
    "Importance: hard work",
    "Importance: having ambition",
    "Importance: wealthy family",
    "Importance: educated parents",
    "Importance: race",
    "Importance: gender",
    "Importance: knowing people",
    "Importance: political connections",
    "Reasons for pay: well job",
    "Reasons for pay: hard job",
    "Reasons for pay: support family",
    "Reasons for pay: has children"
  )

  sum.issp09<- knitr::kable(desc.issp,format =table_format,
                   booktabs= T,
                   align = "l",
                   linesep = "",
                   caption = "Descriptive statistics for the ISSP 2009 sample") %>%
    kableExtra::kable_styling(full_width = T,
                              font_size = 10,
                              position = "center",
                              latex_options = c("HOLD_position"));sum.issp09
```

Figure \@ref(fig:likert-issp) shows the distribution of responses across the selected items in their corresponding dimensions. On the one side, we see that there is a high degree of importance attributed to factors such as hard work and ambition in the process of getting ahead, concentrating 94.9% and 92.8% in the fairly important to essential categories. By contrast, the perception non-meritocratic aspects is rated lower that the meritocratic ones, particularly for gender, race and political connections. Regarding normative preferences, we observe that the meritocratic ones are deemed as important (from fairly to essential) for nearly the whole sample, support that decreases slightly for the non-meritocratic ones. Still, we have to consider that the preferences for non-meritocratic aspects in this case refer to the distributive principle of need rather than personal background (as in the perceptions of non-meritocracy), which certainly is one of the limitations of this set of items for the operationalization of the proposed conceptual framework with this dataset.

@@

```{r likert-issp,fig.height=11,fig.width=11,fig.cap="Distribution of responses in the ISSP meritocracy items", out.width='80%'}

#dev.copy(png,'output/images/Fig2.plotlikert-study1.png') # save image

  if (!require("pacman")) install.packages("pacman")
  pacman::p_load(cowplot, dplyr)
  #Final plot
  cowplot::add_sub(
    cowplot::plot_grid(p1,p2,p3,p4,leg, #plot 1,2,3,4 and legend (bottom)
      align = "hv",
      axis = "a",
      nrow = 5, #5 plots
      ncol = 1, #1 column
      rel_heights = c(1,2,1,1,0.25) #fig size within plot
      ),
    label = paste0(
      "Source: Authors calculation based on International Social Survey Programme - Social Inequality 2009",
      " (n=",dim(likert_issp)[1],")"
      ),        # Caption: source and sample size
    size = 10,  # Caption font size
    hjust = 0.25# Caption align
    ) %>%
    cowplot::ggdraw()
#dev.off()
```

In terms of the association between the indicators, Figure \@ref(fig:corr-issp) shows the polychoric correlation matrix. Firstly, we observe in general that the moderat to high associations are between the pairs of items representing each of the dimensions as described in \@ref(tab:table-issp). The exception here is the dimension of non-meritocratic perception, in which there are six items which appear mostly associated by pairs according to their specific topics (family background, personal background and connections). Still, between this same set of items there are moderate correlations (around 0.3), which could anticipate a single latent factor underlying non meritocratic perception and that is tested next through confirmatory factor analysis.

```{r corr-issp,fig.cap = "Perceptions and preferences for ISSP  meritocracy items' polychoric correlations", fig.align='center',fig.height=6,fig.width=8}

#dev.copy(png,'output/images/Fig3.plotcor-sutdy1.png') # save image

>>>>

---
title: "Measuring Perceptions and Preferences for Meritocracy"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
author: "Juan Carlos Castillo, Julio Iturra, Luis Maldonado, Francisco Meneses & Jorge Atria <br><br>"
abstract: "Economic and social inequalities have generated growing concerns and crises across contemporary societies. One of the mechanisms proposed by social sciences to explain the persistence of inequality is the belief in meritocracy, which would legitimize economic disparities based on differences in effort and talent. Despite its wide use as a concept, empirical research on meritocracy is relatively novel and characterized by diverse conceptualization and measures that make the findings and their interpretation rather inconsistent. Most of the studies in the area have relied upon secondary data to operationalize meritocracy, with a wide variation in the use and interpretation of the same survey items. Taking into account the extant literature that uses measures of meritocracy, this article identifies a series of drawbacks and inconsistencies within and between studies regarding conceptualization and operationalization. Based on this critical analysis, we propose a conceptual framework for studying perceptions and preferences for meritocracy and non-meritocracy, which is then tested through confirmatory analysis using ISSP (International Social Survey Programme) data as well as a novel survey designed specifically with this purpose (N=2,141). Our results support the conceptual framework and its operationalizarion, although with a better fit for the proposed scale than for the ISSP survey. Our discussion highlights the importance of considering different dimensions in order to advance in the study of meritocracy."
output:
  bookdown::word_document2:
    number_sections: false
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
     template: null
     toc: false
     number_sections: false
linkcolor: blue
bibliography:
  - input/bib/meritocracy.bib
csl: input/bib/apa7-singlespace.csl
editor_options:
  chunk_output_type: console
geometry: margin=0.78in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage[figurename=Fig.]{caption}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
always_allow_html: true # for word output
fontsize: 10pt
---

@@

@@

Economic inequality and income concentration have become topics of growing concern over the last years, acquiring even more salience in times of pandemic and economic crises. This has led to a series of social movements in different societies as well as diverse critical analyses regarding the development of capitalism and its consequences [@streeck_politics_2014; @piketty_capital_2014]. In this context, the study of views, preferences, and perceptions of inequality has acquired relevance in the social sciences, in topics such as redistributive preferences [@alesina_fairness_2005; @dimick_models_2018], the legitimization of economic inequality [@schroder_income_2017] and the functioning of meritocracy [@duru-bellat_who_2012; @mijs_paradox_2019; @reynolds_perceptions_2014; @atria_economic_2020]. Within this context, the ideal of meritocracy has been strongly challenged as an unfulfilled promise of modern societies that allows the perpetuation of social inequalities [@sandel_tyranny_2020; @goldthorpe_myth_2003].

In general, meritocracy is defined as a system of distribution of resources and rewards based on individual merit, which in its original conception is a combination of talent and effort [@young_rise_1962]. This traditional conception of merit places in a secondary position the possible interference of structural or non-meritocratic factors, such as inheritance, personal contacts, and luck [@breen_class_1999; @saunders_might_1995; @yair_meritocracy_2007; @land_we_2006; @young_rise_1962]. Social psychology and sociology have studied the characteristics and consequences of beliefs in meritocracy, under the general hypothesis that a greater belief in meritocracy emphasizes the role of the individual over structural factors in personal achievements, leading to greater legitimization of inequalities [@preminger_meritocracy_2020; @trump_when_2020; @hadjar_meritokratie_2008; @madeira_primes_2019]. Such research has raised criticism of meritocracy as a moral standard of distribution given the preponderant weight of non-meritocratic elements upon the individual status and social mobility [@sandel_tyranny_2020; @witteveen_reconsidering_2020; @arrow_meritocracy_2000; @goldthorpe_myth_2003; @markovits_meritocracy_2019; @khan_privilege_2013].

Due to the role that meritocratic beliefs play in the justification of individual achievement (or failure) in contemporary societies [@hadjar_meritokratie_2008; @markovits_meritocracy_2019; @sandel_tyranny_2020], multiple studies have evaluated the relationship between meritocratic beliefs and personal and/or contextual characteristics. For example, some studies have linked meritocracy to the reinforcement of socio-economic, gender, and ethnic stereotypes [@madeira_primes_2019; @girerd_neoliberalism_2020; @girerd_neoliberalism_2020], as well as the effects of meritocratic beliefs in educational [@generett_stories_2020; @owens_engines_2020] and organizational contexts [@perez_advancing_2020; @aiello_new_2019]. Most of such studies so far have used indicators from existing standard social surveys, such as the International Social Survey Programme (ISSP), to measure meritocracy. However, as we will show later, the concepts used as well as the instruments for measuring meritocracy vary extensively among studies. In many cases, similar phenomena are associated with different indicators, and conversely, often different phenomena are measured with similar indicators. Such trends not only limit the comparability of studies, but also the ability to understanding the effects of meritocratic beliefs across different disciplines and lines of research.

@@

In the following, we discuss four critical aspects in studies conceptualizing and measuring meritocracy, based upon which we develop a conceptual and measurement proposal.

@@

One recent definition of meritocracy by @mijs_paradox_2019 is the following: "When I discuss meritocracy beliefs, I am referring to citizens' belief in the importance of hard work relative to structural factors." [@mijs_paradox_2019, pg.9]. In the subsequent operationalization, this is associated with the following question and indicator: "how important you think it is for getting ahead in life: (a) hard work", scored from 1 to 5 on a likert scale. The assumptions behind such definition are worth discussing in light of the conceptual meaning of meritocracy and its possibilities of operationalization.

The item used by @mijs_paradox_2019 is part of an items' battery present in several international surveys, usually called "reasons to get ahead". This battery displays a series of indicators related to what people consider important to get ahead in life: hard work, education, ambition, a wealthy family, the right connections, religion, race, and gender. Therefore, when considering only one of the items, it means that other aspects that could be associated with talent, such as education, would not be deemed as meritocratic. As he points out: "Hard work is arguably the most meritocratic part of Michael Young's equation: 'Merit = Intelligence + Effort', for the simple fact that intelligence itself is influenced by a non-meritocratic factor: who your parents happen to be" (p.5).

In this conceptual and measurement approach of meritocratic beliefs, we can observe a couple of strong conceptual assumptions: a) effort would not depend on parental influence, and b) talent (as innate ability) is not meritocratic (contrary to Michael Youngs' original conceptualization). This conceptual and measurement-based assumption is found in other studies, which also assumes effort to be the main and only aspect of meritocracy [@girerd_neoliberalism_2020; @bubak_perceptions_2019], raising the question: Is effort the only dimension behind the concept of meritocracy? The question of whether talent (as intelligence and ability) is or is not considered meritocratic is certainly an interesting topic to discuss from a philosophical point of view, but for those working on empirical studies, it should face empirical scrutiny. Besides, the consideration of talent as part of meritocracy certainly opens some interesting avenues of research, for instance in studies showing that for the elites meritocracy is more related to talent, whereas effort is something more characteristic of the meritocracy of the middle and low classes [@atria_economic_2020].

@@

Several approaches to the empirical study of meritocracy based on public opinion surveys refer to the concept of _beliefs_, but with wide differences in meanings and operationalizations. To illustrate this point let us consider the most commonly used items for operationalizing meritocratic beliefs in survey research (as ISSP), which is the "reasons to get ahead" battery. It consists of a series of items asking "how important you think it is for getting ahead in life" and then lists several factors, such as effort, education, parental wealth, and contacts. Another version of this same battery used in other surveys - sometimes together with the previous one - asking about "how important you think it _should_ be ...", and then listing the same concepts.  Therefore, the question raised here is: Which one of both is a "belief": what _is_ or what it _should_ be?

The term belief has an ambiguous character in the literature, conceived as "idea-elements" by @converse_nature_1964 or "considerations" by @zaller_nature_1992. As @kluegel_beliefs_1986 pointed out about the scope of beliefs: "This usage encompasses such more specific social-psychological concepts as values, perceptions, and attitudes" (p.30). Therefore, beliefs cover almost anything related to subjective factors. To this regard, a relevant distinction in the field of inequality beliefs was made by @janmaat_subjective_2013: "Perceptions refer to subjective estimates of existing inequality (i.e. thoughts about what is). Beliefs are here defined as normative ideas about just inequality (i.e. thoughts about what should be)"(p.359). Several papers dealing with meritocracy use the term beliefs (i.e. what should be), while actually referring to perceptions (i.e. what is). This occurs for instance in @reynolds_perceptions_2014, in which the term belief is used to talk about what @janmaat_subjective_2013 refers to as perceptions, whereas other authors use general terms such as attitudes [@kunovich_systems_2007]. The first attempt to shed light on this issue in meritocracy research was made by @duru-bellat_who_2012, who used the question "how important should the number of years spent in education and training be in deciding how much money people ought to earn?" as a proxy for "desired" meritocracy (normative beliefs). They then determined "perceived" meritocracy, using the questions: "Would you say that in your country, people are rewarded for their efforts?" and "... people are rewarded for their skills?".

Is the belief in meritocracy a perception or a preference with normative meaning? In order to expand the analytical conceptual framework, we believe that both dimensions should be included as proposed by @duru-bellat_who_2012. This opens up the possibility of analyzing whether perceptions and preferences are actually related (i.e. have a high correlation), or whether they are independent aspects of the same phenomenon. As @sonhing_merit_2011 has pointed out, "People can believe that outcomes ought to be distributed on the basis of merit and yet vary in their perceptions of whether this is how society currently operates" (p. 435). In other words, normative beliefs should be considered while taking perceptions into account: a strong normative belief in meritocracy may mean something totally different to someone perceiving high meritocracy than to someone perceiving low meritocracy. To avoid the confusion generated by the term "belief", we propose the terms meritocratic preferences ("what should be"), and meritocratic perceptions ("what is"), as they better reflect the two facets of meritocracy under scrutiny [@castillo_meritocracia_2019].

@@

Some research in meritocracy considers aspects usually opposed to effort and talent for personal achievement, as for instance the use of personal advantages (as contacts or having a wealthy family) to get ahead in life. For instance, @kunovich_systems_2007 used an items' battery listing a number of factors in relation to "How important each should be in deciding pay..." (as @duru-bellat_who_2012 for desired meritocracy). They consider factors such as education and responsibility as meritocratic, giving them a value of 1 is considered "essential" in the scale response, whereas factors such as having a family or children are valued 1 when rated as "not important at all" (i.e. reverse coded). The assumption behind this approach is that rejecting a supposed non-meritocratic aspect (as having family and/or children) implies a stronger belief in meritocracy. A similar approach of reverse-coding non-meritocratic items was taken by @newman_false_2015, using the same principles applied in the "Preference for the Merit Principle Scale" [@davey_preference_1999].

@@

The columns "Perceptions" and "Preferences" represent the distinction between these two concepts, usually confused under the label "beliefs" [@castillo_meritocracia_2019]. Perceptions refer to the extent to which people observe that meritocracy functions or apply in their society, which relates to items such as "Hard work is important to get ahead in society". Preferences refer to normative expectations that are usually linked to a "should" expression (e.g. whether hard work should be related to payment). The rows in Figure \@ref(fig:merit-model) consider the distinction between meritocratic and non-meritocratic dimensions [@reynolds_perceptions_2014], usually treated as different ends of the same continuum in previous research. Non-meritocratic elements refer to the use of resources such as personal contacts or family advantages to getting ahead in life.

@@

The analysis is organized into three studies. Study 1 serves as a background to our instrument proposal by attempting to operationalize the meritocracy dimensions with available secondary data from the inequality module of the International Social Survey Programme (ISSP). Although we are aware that the available indicators in this dataset do not allow a precise operationalization of the conceptual model, we know that it is the most widely used data in comparative meritocracy studies (as well as overall subjective inequality) and therefore it is certainly useful to look at its potentials and limitations to this regard. Study 2 is the core of this paper and consists of the proposal of a new scale for measuring meritocratic perceptions and preferences. Finally, in Study 3 we perform an additional validity analysis of the meritocracy scale.

@@

The data corresponds to the last available wave of the social inequality module from the International Social Survey Programme (ISSP), which is the most specialized international comparative survey in perceptions, attitudes, and beliefs about inequality-related issues. This wave corresponds to the year 2009 and covers attitudes towards a series of topics dealing with social inequality across 41 countries and 56021 individuals. Although there is data available from this module for previous waves (1987, 1992, and 1999), unfortunately, there are several variables that are important for this research which were not included, particularly in 1987 and 1999, the reason why we only use the 2009 wave (during the development of this research the 2019 wave became available, but so far only with a restricted set of countries).

@@

- _Preferences for meritocracy-non meritocracy_: for the operationalization of normative preferences we used a list of items related to reasons for pay battery. The ISSP question was: _In deciding how much people ought to earn, how important should each of these things be, in your opinion_, rated in the same _essential-non important at all_ scale (1 to 5) as the questions for meritocratic perceptions.

@@

The estimation was performed using Confirmatory Factor Analysis (CFA). CFA was conducted using the `lavaan` R package (version 0.6-3; @rosseel_lavaan_2012), with diagonally weighted least squares (DWLS) estimation due to the items' ordinal level of measurement [@kline_principles_2016; @rosseel_lavaan_2012]. As recommended by @brown_confirmatory_2015, we assessed model fit by jointly considering the comparative fit index and Tucker-Lewis Index (CFI and TLI; acceptable fit > 0.95), Root of the average squared residual approximation (RMSEA; acceptable fit < 0.08),  Chi-square: (p-value; acceptable fit > 0.05, and Chi-square ratio > 3).

@@

Figure \@ref(fig:likert-issp) shows the distribution of responses across the selected items in their corresponding dimensions. On the one side, we see that there is a high degree of importance attributed to factors such as hard work and ambition in the process of getting ahead, concentrating 94.9% and 92.8% in the fairly important to essential categories. By contrast, the perception of non-meritocratic aspects is rated lower than the meritocratic ones, particularly for gender, race, and political connections. Regarding normative preferences, we observe that the meritocratic ones are deemed as important (from fairly to essential) for nearly the whole sample, support that decreases slightly for the non-meritocratic ones. Still, we have to consider that the preferences for non-meritocratic aspects, in this case, refer to the distributive principle of need rather than personal background (as in the perceptions of non-meritocracy), which certainly is one of the limitations of this set of items for the operationalization of the proposed conceptual framework with this dataset.

@@

```{r likert-issp,fig.height=11,fig.width=11,fig.cap="Distribution of responses in the ISSP meritocracy items", out.width='80%'}
  # dev.copy(png,'output/images/Fig2.plotlikert-study1.png', width = 800, height = 600) # save image
  if (!require("pacman")) install.packages("pacman")
  pacman::p_load(cowplot, dplyr)
  #Final plot
  cowplot::add_sub(
    cowplot::plot_grid(p1,p2,p3,p4,leg, #plot 1,2,3,4 and legend (bottom)
      align = "hv",
      axis = "a",
      nrow = 5, #5 plots
      ncol = 1, #1 column
      rel_heights = c(1,2,1,1,0.25) #fig size within plot
      ),
    label = paste0(
      "Source: Authors calculation based on International Social Survey Programme - Social Inequality 2009",
      " (n=",dim(likert_issp)[1],")"
      ),        # Caption: source and sample size
    size = 10,  # Caption font size
    hjust = 0.25# Caption align
    ) %>%
    cowplot::ggdraw()
    # dev.off()
```

In terms of the association between the indicators, Figure \@ref(fig:corr-issp) shows the polychoric correlation matrix. Firstly, we observe in general that the moderate to high associations are between the pairs of items representing each of the dimensions as described in \@ref(tab:table-issp). The exception here is the dimension of non-meritocratic perception, in which there are six items that appear mostly associated by pairs according to their specific topics (family background, personal background, and connections). Still, between this same set of items, there are moderate correlations (around 0.3) which could anticipate a single latent factor underlying non-meritocratic perception and that is tested next through confirmatory factor analysis.

```{r corr-issp,fig.cap = "Perceptions and preferences for ISSP  meritocracy items' polychoric correlations", fig.align='center',fig.height=6,fig.width=8}
  # dev.copy(png,'output/images/Fig3.plotcor-study1.png', width = 800, height = 600) # save image
  load(file = "input/data/proc/data_s1.RData")
  if (!require("pacman")) install.packages("pacman")
  pacman::p_load(dplyr, corrplot, lavaan)
  # create polychoric correlation matrix
  cor_issp<-
    data_s1 %>%
    dplyr::select(hwork,ambition,wealthy,pareduc,
                  race,gender,people,polcone,
                  welljob,hardjob,family,child) %>%
    lavaan::lavCor(., ordered=names(.))
  diag(cor_issp) = NA #set diagonal values to NA
  # Set Row names of the matrix
  rownames(cor_issp) <- c("A. Hard work",
                         "B. Having ambition",
                         "C. Wealthy family",
                         "D. Educated parents",
                         "E. Race",
                         "F. Gender",
                         "G. Knowing people",
                         "H. Political connections",
                         "I. Well job",
                         "J. Hard job",
                         "K. Support family",
                         "L. Has children")
  #set Column names of the matrix
  colnames(cor_issp) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)",
                         "(H)", "(I)","(J)","(K)","(L)")
